---
title: "Exercise 04"
author: "Julia Frank"
date: "2025-03-14"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
})
```

## Download Data

**Find and download from ENCODE the peaks (i.e. bed-like format) for the
following in mouse embryonic stem cells (mESC) :**

-   **Transcription factor p300**

<!-- -->

-   **Histone modifications H3K4me3, H3K4me1, H3K27ac, and H3K27me3**

<!-- -->

-   **(when there are replicates, we recommend using the bed file
    denoted as “conservative IDR thresholded peaks”)**

Since I only found data for the transcription factor p300 in the
ES-Bruce4 strain, I decided to use it for all the data to ensure
comparability between the transcription factor and the histone
modifications. All the data is also from the same lab.

For the EP300 ChIP-seq data, I used the latest conservative
IDR-thresholded peaks, while for the Histone ChIP-seq data, I used the
latest default peaks (typically pseudoreplicated peaks for two isogenic
replicates).

ENCODE provided warnings about data quality for all datasets, such as
insufficient read depth.

```{r}
options(timeout=3600) # Increase the download timeout
dir.create("raw_data") # Create a directory for the data

# Transcription factor
# p 300 (https://www.encodeproject.org/experiments/ENCSR000CCD/)
download.file("https://www.encodeproject.org/files/ENCFF653YVR/@@download/ENCFF653YVR.bed.gz", "raw_data/p300.bed.gz", mode="wb")
peaks_p300 <- rtracklayer::import("raw_data/p300.bed.gz", format="narrowPeak")

# Histone modifications
# H3K4me3 (https://www.encodeproject.org/experiments/ENCSR000CBG/)
download.file("https://www.encodeproject.org/files/ENCFF974BMC/@@download/ENCFF974BMC.bed.gz", "raw_data/H3K4me3.bed.gz", mode="wb")
peaks_H3K4me3 <- rtracklayer::import("raw_data/H3K4me3.bed.gz", format="narrowPeak")

# H3K4me1 (https://www.encodeproject.org/experiments/ENCSR000CBF/)
download.file("https://www.encodeproject.org/files/ENCFF426IIV/@@download/ENCFF426IIV.bed.gz", "raw_data/H3K4me1.bed.gz", mode="wb")
peaks_H3K4me1 <- rtracklayer::import("raw_data/H3K4me1.bed.gz", format="narrowPeak")

# H3K27ac (https://www.encodeproject.org/experiments/ENCSR000CDE/)
download.file("https://www.encodeproject.org/files/ENCFF274UIB/@@download/ENCFF274UIB.bed.gz", "raw_data/H3K27ac.bed.gz", mode="wb")
peaks_H3K27ac <- rtracklayer::import("raw_data/H3K27ac.bed.gz", format="narrowPeak")

# H3K27me3 (https://www.encodeproject.org/experiments/ENCSR000CFN/)
download.file("https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz", "raw_data/H3K27me3.bed.gz", mode="wb")
peaks_H3K27me3 <- rtracklayer::import("raw_data/H3K27me3.bed.gz", format="narrowPeak")
```

## Question

**Of the p300 peaks, what proportion overlap each of the marks?
(expected form of the answer: one % for each of the marks)**

```{r}
# function to find proportion of overlaps
overlaps <- function(p300, histone){
  o <- findOverlaps(p300, histone, type='any')
  n_p300 <- length(peaks_p300)
  prop <- length(o)/n_p300 * 100
  return(prop)
}

# Compute the proportion of p300 peaks overlapping each mark
p300_H3K4me3 <- overlaps(peaks_p300, peaks_H3K4me3)
p300_H3K4me1 <- overlaps(peaks_p300, peaks_H3K4me1)
p300_H3K27ac <- overlaps(peaks_p300, peaks_H3K27ac)
p300_H3K27me3 <- overlaps(peaks_p300, peaks_H3K27me3)

# Print results
cat("Proportion of p300 peaks overlapping each of the marks:")
cat("- H3K4me3:", round(p300_H3K4me3,2), "%")
cat("- H3K4me1:", round(p300_H3K4me1,2), "%")
cat("- H3K27ac:", round(p300_H3K27ac,2), "%")
cat("- H3K27me3:", round(p300_H3K27me3,2), "%")

```
